---
title: "heatmap4 Vignette"
author: "Rohan Bhanot"
date: "10/31/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction
This package takes the original heatmap function and reduces the arguement complexity. 

A heat map is a false color image (basically image(t(x))) with a dendrogram added to the left side and to the top. Typically, reordering of the rows and columns according to some set of values (row or column means) within the restrictions imposed by the dendrogram is carried out.

One package that is used is Bioconductor. 
The simplest way to install it is:

```{r}
if (F) {
  ## Calling Bioconductor
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

  BiocManager::install("marray")
  load("data.RData")
}
```

## 2. Loading Libraries

The libraries to call are the following:

```{r}
library(RColorBrewer)
library(marray)
```

```{r include=FALSE}
## Wrapped heatmap function
generate_heatmap <- function(x, col_lab = c(TRUE, FALSE), row_lab = c(TRUE, FALSE), col_lab_vtr = NULL,
                             row_lab_vtr = NULL, col_anno = c(TRUE, FALSE), row_anno = c(TRUE, FALSE), col_info = NULL,
                             row_info = NULL, col_anno_var = NULL, row_anno_var = NULL, col_var_info = NULL,
                             row_var_info = NULL, col_dend = c(TRUE, FALSE), row_dend = c(TRUE, FALSE),
                             # review col/row_clust and _dend
                             col_clust = NULL, row_clust = NULL,
                             plot_info = c("sideLabCol" = NULL, "sideLabRow" = NULL, "cexColSide" = NULL, "cexRowSide" = NULL),
                             file_name = NULL, h_title = NULL, ...)
{

  # ## Row and Column Names
  # row_new_lab <- rownames(x)
  # col_new_lab <- colnames(x)

  #--------------------------------------------------------------------------------------------
  ## Annotations
  row_anno <- row_anno[1]

  if (row_anno) {

    # Checks to see if row_info is a valid input
    if (class(row_info) != 'data.frame') {
      stop("row_info is meant to be a data.frame")
    }

    # Defining row annotation variable selection
    row_var <- c()

    # Looping through row annotation data frame
    if (is.null(row_anno_var)) {
      for (r in 1:ncol(row_info)) {
        if (sum(!duplicated(row_info[ ,r])) < nrow(row_info)) {
          row_var <- c(row_var, colnames(row_info)[r])
        }
      }
    } else {
      k <- match(colnames(row_info), row_anno_var)
      k <- k[!is.na(k)]
      if (length(k) == 0) {
        row_var <- colnames(row_info)
        cat("Selection of row annotation variables does not match column names in original data frame.\n")
      } else {
        row_var <- row_anno_var[k]
        if (length(k) != length(row_anno_var)) cat("Selection of column annotation variables does not match column names in original data frame.\n")
      }


    }

    color_vec_default <- c("skyblue", "blue", "yellow", "purple", "black", "red", "orange", "green", "cyan", "darkgreen")

    row_color <- matrix(nrow = length(row_var), ncol = nrow(row_info))
    rownames(row_color) <- row_var

    for (v in 1:length(row_var)) {
      if (is.null(row_var_info)) {
        color_vec <- color_vec_default
      } else {
        if (row_var[v] %in% names(row_var_info)) {
          if ("color" %in% names(row_var_info[[row_var[v]]])){
            color_vec <-  row_var_info[[row_var[v]]]$color
          }
        }
      }

    if (is.numeric(row_info[ ,row_var[v]]) & length(unique(row_info[ ,row_var[v]])) > 5) {

        varib <- row_info[ ,row_var[v]]
        varib <- varib - min(varib, na.rm=T) + 1
        varib <- varib / min(varib, na.rm=T)

        varib <- round(varib)
        lim <- range(varib,na.rm=T)

        grpUniq <- lim[1]:lim[2]
        rowColUniq <- maPalette(high=colList2[2], low=colList2[1], k=length(grpUniq))


        row_color[v, ] <- rowColUniq[varib]

      } else {
        dat <- row_info[ ,row_var[v]]
        datUniq <- sort(unique(dat))
        for (v2 in 1:length(datUniq)) {
          j <-  which(dat == datUniq[v2])
          row_color[v, j] <- color_vec[v2]

        }
      }
    }
    RowSideColors <- row_color
  } else {
    RowSideColors <- NULL
  }
  #------------------------------------------
  col_anno <- col_anno[1]

  if (col_anno) {

    # Checks to see if col_info is a valid input
    if (class(col_info) != 'data.frame') {
      stop("col_info is meant to be a data.frame")
    }

    # Defining column annotation variable selection
    col_var <- c()

    # Looping through column annotation data frame
    if (is.null(col_anno_var)) {
      for (r in 1:ncol(col_info)) {
        if (sum(!duplicated(col_info[ ,r])) < nrow(col_info)) {
          col_var <- c(col_var, colnames(col_info)[r])
        }
      }
    } else {
      k=match(colnames(col_info),col_anno_var)
      k=k[!is.na(k)]
      if (length(k)==0) {
        col_var=colnames(col_info)
        cat("Selection of column annotation variables does not match column names in original data frame.\n")
      } else {
        col_var=col_anno_var[k]
        if (length(k)!=length(col_anno_var)) cat("Selection of column annotation variables does not match column names in original data frame.\n")
      }

    }

    color_vec_default <- c("skyblue", "blue", "yellow", "purple", "black", "red", "orange", "green", "cyan", "darkgreen")

    col_color <- matrix(nrow = length(col_var), ncol = nrow(col_info))
    rownames(col_color) <- col_var

    for (v in 1:length(col_var)) {
      if (is.null(col_var_info)) {
        if (is.numeric(col_info[ ,col_var[v]]) & length(unique(col_info[ ,col_var[v]])) > 5) {
          color_vec <- c("white","black")
        } else {
          color_vec <- color_vec_default
        }
      } else {
        if (col_var[v] %in% names(col_var_info)) {
          if ("color" %in% names(col_var_info[[col_var[v]]])) {
            color_vec <-  col_var_info[[col_var[v]]]$color
          }
        }
      }

      if (is.numeric(col_info[ ,col_var[v]]) & length(unique(col_info[ ,col_var[v]])) > 5) {

        varib <- col_info[ ,col_var[v]]
        varib = varib - min(varib, na.rm=T) + 1
        varib = varib / min(varib, na.rm=T)

        varib = round(varib)
        lim = range(varib,na.rm=T)

        grpUniq=lim[1]:lim[2]
        colColUniq=maPalette(high=colList2[2],low=colList2[1],k=length(grpUniq))


        col_color[v, ] <- colColUniq[varib]

      } else {
        dat <- col_info[ ,col_var[v]]
        datUniq <- sort(unique(dat))
        if (length(datUniq) > length(color_vec)) {
          color_vec <- rainbow(length(datUniq))
          cat("Not enough colors for ,",col_var[v],"; They have been assigned random colors in the meantime.\n")
        }
        for (v2 in 1:length(datUniq)) {
          j <-  which(dat == datUniq[v2])
          col_color[v,j] <- color_vec[v2]

        }
      }
    }
    ColSideColors <- col_color
  } else {
    ColSideColors <- NULL
  }
  #--------------------------------------------------------------------------------------------
  # Global Plot Info
  nc <- ncol(x)
  nr <- nrow(x)

  cexColSide <-  1
  cexRowSide <-  1
  cexCol <- 0.2 + 1 / log10(nc)
  cexRow <- 0.2 + 1 / log10(nr)

  if (!(is.null(plot_info))) {

    if ("cexRowSide" %in% names(plot_info)) {
      cexRowSide <- plot_info$cexRowSide
    }
    if ("cexColSide" %in% names(plot_info)) {
      cexColSide <- plot_info$cexColSide
    }
    if ("cexCol" %in% names(plot_info)) {
      cexCol <- plot_info$cexCol
    }
    if ("cexRow" %in% names(plot_info)){
      cexRow <- plot_info$cexRow
    }
  }

  #--------------------------------------------------------------------------------------------
  ## Labels
  row_lab <- row_lab[1]

  if (row_lab) {
    if(!(is.null(row_lab_vtr))) {
      labRow <- row_lab_vtr
    } else {
      labRow <- rownames(x)
    }
  }
  else {
    labRow <- NA
  }
  #------------------------------------------
  col_lab <- col_lab[1]

  if (col_lab) {
    if(!(is.null(col_lab_vtr))){
      labCol = col_lab_vtr
    } else {
      labCol <- colnames(x)
    }
  }
  else {
    labCol <- NA
  }
  #--------------------------------------------------------------------------------------------
  ## Dendograms
  row_dend <- row_dend[1]

  if (row_dend) {
    Rowv <- row_clust
  }
  else {
    Rowv <- NA
  }
  #------------------------------------------
  col_dend <- col_dend[1]

  if (col_dend) {
    if (is.null(col_clust)) { # | class(col_clust %in% c("dendogram", "hclust"))
      Colv <- col_clust
    }
    else {
      stop("Error in column clustering class.")
    }
  }
  else {
    Colv <- NA
  }
  #--------------------------------------------------------------------------------------------
  ## Output Files
  if (!is.null(file_name)){

    element <- strsplit(file_name, ".", fixed = TRUE)

    if (element[[1]][2] == "pdf") {
      pdf(file_name, paper = "letter")
    }
    else if (element[[1]][2] == "jpeg") {
      jpeg(filename = file_name)
    }
    else if (element[[1]][2] == "png") {
      png(filename = file_name)
    }
    else if (element[[1]][2] == "tiff") {
      tiff(filename = file_name)
    }
    else {
      stop("File name not in valid format: filename.type")
    }
  }
  #--------------------------------------------------------------------------------------------
  ## Heatmap Output
  heatmap4(x = x, Rowv = Rowv, Colv = Colv, distfun = dist,  symm = FALSE,
           ColSideColors = ColSideColors, RowSideColors = RowSideColors, labCol = labCol, labRow = labRow,
           scale = "none", na.rm = FALSE, margins = c(5, 5), main = h_title, xlab = NULL, ylab = NULL, zlm = limit,
           high = cols[1], low = cols[2], mid = cols[3], cexRowSide = cexRowSide, cexColSide = cexColSide, cexRow = cexRow,
           cexCol = cexCol, ...)

  #--------------------------------------------------------------------------------------------
  ## Clears graphics on device
  dev.off()
}
```

## 3. Example Data

We will use the preloaded mtcars data set to map and randonly generate column and row annotations.
```{r}
# Main numeric matrix to heatmap
test_x <- as.matrix(mtcars)

# Column annotation df
set.seed(827)

col_rows <- ncol(mtcars)
col_columns <- 3

column_data <- rnorm(col_rows * col_columns)
column_df <- matrix(data = column_data, nrow = col_rows, ncol = col_columns)
colnames(column_df) <- c('C1', 'C2', 'C3')
column_df <- as.data.frame(column_df)

# Row annotation df
set.seed(957)

row_rows <- nrow(mtcars)
row_columns <- 2

row_data <- rnorm(row_rows * row_columns)
row_df <- matrix(data = row_data, nrow = row_rows, ncol = row_columns)
colnames(row_df) <- c('R1', 'R2')
row_df <- as.data.frame(row_df)

```

## 4. Use Cases

### 4.1 Annotations and Dendrograms

To display the heatmap with just column annotations and dendrograms:
* NOTE: This is the simplest way to call the heatmap4 function. Defaults are accounted for, see documentation for full list of   defualt options for each arguement.

```{r}

generate_heatmap(test_x, row_anno = FALSE, col_info = column_df,  row_dend = FALSE)

```








Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
