---
title: "heatmap4 Vignette"
author: "Rohan Bhanot"
date: "10/31/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction
This package takes the original heatmap function and reduces the arguement complexity. 

A heat map is a false color image (basically image(t(x))) with a dendrogram added to the left side and to the top. Typically, reordering of the rows and columns according to some set of values (row or column means) within the restrictions imposed by the dendrogram is carried out.

One package that is used is Bioconductor. 
The simplest way to install it is:

```{r}
if (F) {
  if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

  BiocManager::install("marray")
  load("data.RData")
}
```

## 2. Loading Libraries

The libraries to call are the following:

```{r}
library(RColorBrewer)
library(marray)
```

## 3. Example Data

We will use the preloaded mtcars data set to map and randonly generate column and row annotations.
```{r}
# Main numeric matrix to heatmap
x <- as.matrix(mtcars)

# Column annotation df
set.seed(827)

anno_rows <- ncol(mtcars)
anno_columns <- 3

random_data <- rnorm(anno_rows * anno_columns)
annotation_df <- matrix(data = random_data, nrow = anno_rows, ncol = anno_columns)
colnames(annotation_df) <- c('C1', 'C2', 'C3')
annotation_df <- as.data.frame(annotation_df)



```

## 4. Use Cases

### 4.1 Basic 








Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
